---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The `dplyr` package is being used here to process the data better. 
`lubridate` is for easier date manipulation and `lattice` for the graph. 
``` {r tools, results = "hide"}
library(dplyr)
library(lubridate)
library(lattice)
```

The data is then read from the csv file. 
``` {r readData}
activity = read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
`dplyr` has a `group_by` function which can be used to find the total number 
of steps taken per day. The observations which don't have a steps are removed
for this. 

``` {r}
steps_by_date <- activity %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))
```

The `steps_by_date` variable produced in the previous step is then used to 
plot a histogram. 

``` {r histogram_plot1}
hist(steps_by_date$total_steps, col = 'red', xlab = "Total steps each day", 
     main = "Steps per day")
```

The mean and median are taken on the total_steps column
``` {r}
mean(steps_by_date$total_steps)
median(steps_by_date$total_steps)
```

## What is the average daily activity pattern?
Finding the average steps taken across all intervals :

``` {r}
steps_by_interval <- activity %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(avg_steps = mean(steps))
```

Plot a time series plot of the 5 minute interval and the average number of steps.


```{r interval_steps_plot2}
plot(steps_by_interval$interval, steps_by_interval$avg_steps, type = "l",
     col = "blue", xlab = "Interval", ylab = "Avg Steps", 
     main = "Avg. steps taken per interval")
```

Filtering by the interval which would have the maximum number of steps. 
This is then printed out to give the interval (along with the avg steps.)
``` {r}
max_interval <- steps_by_interval %>%
  filter(avg_steps == max(avg_steps))
max_interval
```

## Imputing missing values
Calculating the rows which have missing values for steps.

``` {r}
empty_steps <- activity %>%
  filter(is.na(steps))
nrow(empty_steps)
```

To do the imputing, 
* First separate out the values which are complete.
* Next calculate the average by interval
* Join this with the empty steps by interval. 
* Mutate empty steps to make steps the average steps by interval
* Remove the average steps column. 

``` {r}
complete_steps <- activity %>%
  filter(!is.na(steps))
complete_steps_avg <- complete_steps %>%
  group_by(interval) %>%
  summarize(avg_steps = mean(steps))
empty_steps <- empty_steps %>%
  inner_join(complete_steps_avg, by = "interval") %>%
  mutate(steps = avg_steps) %>%
  select(-avg_steps)
complete_activity <- bind_rows(complete_steps, empty_steps)
```

Now calculating the steps by day again in the merged result and drawing the 
historgram. 
``` {r imputed_steps_plot3}
complete_steps_day <- complete_activity %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))
hist(complete_steps_day$total_steps, col = 'red', xlab = "Total steps each day", 
     main = "Imputed Steps per day")
mean(complete_steps_day$total_steps)
median(complete_steps_day$total_steps)
```

We see that the mean has remained the same. 

To find the difference in total steps, the difference from the `complete_steps`
and `complete_activity` are calculated
``` {r}
sum(complete_activity$steps) - sum(complete_steps$steps)
```

## Are there differences in activity patterns between weekdays and weekends?
To answer this, first date which was read as factors is converted into character.
Next a column called weekday is added to denote whether the date is weekday or weekend.
To do this we use the `wday` function from `lubridate` package. 
```{r}
complete_activity$date <- as.character(complete_activity$date)
complete_activity <- complete_activity %>%
  mutate(weekday = as.factor(ifelse(wday(date, label = T) %in% c('Sat', 'Sun'), "weekend", "weekday")))
```

Next we find the avergae steps grouped by interval and weekdays. This is then used
for an xyplot with the lattice plotting system. 

``` {r interval_weekday_panels_plot4}
interval_weekday_average <- complete_activity %>%
  group_by(interval, weekday) %>%
  summarize(avg_steps = mean(steps))
xyplot(avg_steps ~ interval | weekday, data = interval_weekday_average, 
       layout = c(1, 2), type = "l")
```

We see from the graph above that there are spikes in the weekdays for certain
intervals whereas in the weekends data is more spread out. 
